Extended Data

### 2022-02-21 doc/manuscript/Dow_Extended_Data.Rmd
remove use of growth "phenology"

## table captions
species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and temperature senisitivity of ring- and diffuse- porous species groups at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest."
tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations of ring width index to monthly $T_{max}$ in univariate and multivariate analyses."

@@

![**Extended Data Figure 4 | Response of stem growth phenology and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`). Shown are relationships between mean $T_{max}$ over the CTW and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. 95% credible intervals are represented by bands. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective CTW. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }

@@

![**Extended Data Figure 5 | Response of stem growth phenology and rates to April mean maximum temperature ($T_{max}$) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are relationships between April $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. 95% credible intervals are represented by bands. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }

>>>>

## table captions
species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the seasonality and temperature senisitivity of ring- and diffuse- porous species groups at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest."
tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring chronologies analyzed and number of significant (at p=0.05) positive or negative correlations of ring width index to monthly $T_{max}$ in univariate and multivariate analyses."

@@

![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (CTW) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** CTW was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`). Shown are relationships between mean $T_{max}$ over the CTW and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. 95% credible intervals are represented by bands. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective CTW. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }

@@

![**Extended Data Figure 5 | Response of stem growth timing and rates to April mean maximum temperature ($T_{max}$) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are relationships between April $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. 95% credible intervals are represented by bands. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }

### 2022-02-21 doc/manuscript/Dow_manuscript_Nature.Rmd
remove use of growth "phenology"

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in growth phenology[@dorangeville_peak_2021; @tumajer_limitation_2022]. 
These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

Growth milestones for both canopy foliage phenology and stem growth occurred 6-10 days earlier, on average, at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig. `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`). 
Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).

![**Figure 2 | Foliage (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each forest plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003). Mean $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`. Mean temperature was calculated for each xylem architecture/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. Leaf phenology years are colored according to temperature in the critical $T_{max}$ window which had the best correlation with timing of greenup (ring-porous window at HF and diffuse-porous at SCBI).](tables_figures/DOYtiming_allyears.png){ width=25% }

@@

Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and C cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the phenology of forest C cycling.
If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_[@churkina_spatial_2005].
Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.

@@

<!--Hypothesis testing -->
Correlation between the dendrometer band-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig. `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
Mixed effects models were used to test the response of growth phenology variables to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the CTW, along with random effects of species and of individual tree. 
We ran separate models for each species group at each site, and for the response of all growth phenology variables to $T_{max}$ (or $T_{min}$). 
This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020; @stan_development_team_stan_2019]. 
In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
<!--
More information on rstanarm's default weakly informative priors:
https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
-->

>>>>

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- (69 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (54 trees at SCBI and 202 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021; @tumajer_limitation_2022]. 
These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

Seaonal milestones for both canopy foliage phenology and stem growth occurred 6-10 days earlier, on average, at SCBI than at Harvard Forest (Fig. `r DOY_timing`, `r  EDT_phenology_parameters`).
Consistent with the results of Ref[@dorangeville_peak_2021], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively; Fig. `r DOY_timing`, `r  EDF_growth_curves_all`, `r  EDT_phenology_parameters`). 
Peak growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species (`r  EDT_phenology_parameters`).

![**Figure 2 | Responses of foliage phenology (a,b) and stem growth timing (c,d) to spring temperatures at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d).** Panels (a-b) show ecosystem-level canopy foliage phenology from 2001-2018, obtained from the MODIS Global Vegetation Phenology product (MCD12Q2.006) for a single pixel at the center of each forest plot, where G = Greenup, M=Mid-greenup, P=peak, and S=Senescence (i.e., beginning of green-down). Panels (c-d) show the dates at which stem growth milestones were achieved, on average, for sampled populations of ring-porous and diffuse-porous trees at SCBI (2011-2020) and Harvard Forest (1998-2003). Mean $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ were estimated using the Bayesian model visualized, with confidence intervals, in `r EDF_pheno_Tmax_climwin`. Mean temperature was calculated for each xylem architecture/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. Leaf phenology years are colored according to temperature in the critical $T_{max}$ window which had the best correlation with timing of greenup (ring-porous window at HF and diffuse-porous at SCBI).](tables_figures/DOYtiming_allyears.png){ width=25% }

@@

Thus, a distinction between interannual variation and directional change may be critical when considering how directional climate change is likely to affect tree growth and ecosystem C dynamics. 
As discussed above, temporal lags between C uptake and release imply that the full effects of warm spring temperatures on forest woody productivity and C cycling are unlikely to be apparent in analyses of interannual variation (including this analysis)[@kannenberg_linking_2019].
Moreover, acclimation of trees to warming temperatures[@gessler_way_2020] and, on longer time scales, species adaptations and shifts in community composition[@fisichelli_temperate_2014] are likely to alter the seasonality of forest C cycling.
If we look across spatial gradients where the latter have had time to play out, we see that warmer spring temperatures are associated with earlier leaf-out[@delgado_differences_2020] and longer growing seasons, which in turn are correlated with greater tree growth[@anderson-teixeira_joint_2021], woody productivity[@banburymorgan_global_2021], and _NEE_[@churkina_spatial_2005].
Thus, warming spring temperatures are expected to increase the biophysical potential for annual tree growth, but that potential is not being realized on an interannual time frame.

@@

<!--Hypothesis testing -->
Correlation between the dendrometer band-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig. `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the CTW, along with random effects of species and of individual tree. 
We ran separate models for each species group at each site, and for the response of all growth parameters to $T_{max}$ (or $T_{min}$). 
This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020; @stan_development_team_stan_2019]. 
In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
<!--
More information on rstanarm's default weakly informative priors:
https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
-->
